---
output:
  html_document: default
  pdf_document: default
  word_document: default
---
EDA with White wine Quality Dataset by VISHAL VIKRAM SINGH
========================================================

#INTRODUCTION

This dataset is related to the white wine of Portuguese version of "Vinho Verde".Here we will apply Exploratory data analysis in order to find out any influence of the chemical properties of wine over the quality of wine which is ranked from 0(very bad) to 9 (very good).

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using
# in your analysis in this code chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk.
# This prevents the code from displaying in the knitted HTML output.
# You should set echo=FALSE for all code chunks in your file.

library(ggplot2)
library(gridExtra)
library(dplyr)
library(corrplot)
library(RColorBrewer)
```

Summary of Data 

```{r echo=FALSE, Load_the_Data}

wine<-read.csv("wineQualityWhites.csv")
dim(wine)


```
Our dataframe consist of 4896 rows and 13 columns.
```{r echo=FALSE,message=FALSE,warning=FALSE}
str(wine)
```

```{r echo=FALSE,message=FALSE,warning=FALSE}
summary(wine)
```

We can observe that median quality of wine is 6 which means 50% of wines are rated more than 6 on the scale of 0 to 9 .It can also be observed that minimum value of citric acid is 0 , which means some wines have no citric Acid. It can also be inferred that some wines have high total sulphur dioxide content.


#DISTRIBUTION OF SINGLE VARIABLES

##1.QUALITY

```{r echo=FALSE, Univariate_Plots}

ggplot(aes(x=quality),data = wine)+geom_bar(fill="orange")+
  scale_x_continuous(breaks = c(3,4,5,6,7,8,9))
```

```{r echo=FALSE,warning=FALSE,message=FALSE}
summary(wine$quality)

```

We can see that most of the wines are of quality rating 6. Only 5 of wines are rated 9. We have very small number of top quality rated wines in the dataset.

##2.ALCOHOL

```{r echo=FALSE, warning=FALSE}
ggplot(aes(x=alcohol),data=wine)+
  geom_histogram(fill="orange",binwidth =0.30)+
  scale_x_continuous(breaks = seq(8,15,0.50))

```

```{r echo=FALSE,warning=FALSE}
summary(wine$alcohol)

```

Aevrage alcohol quanity across the wine dataset is 10.51 (percentage by volume).It can be observed that there is peak around 9.3 . The distribution is roughly normal.

##3.RESIDUAL SUGAR 

```{r echo=FALSE,warning=FALSE}
ggplot(aes(x=(residual.sugar)),data=wine)+
  geom_histogram(fill="orange",binwidth = 0.1)+
  xlim(0,quantile(wine$residual.sugar,0.99))
```

Here we can see that distribution of residual sugar across the wine dataset is skewed . We can apply some transformations to investigate it further.

```{r echo=FALSE, warning=FALSE,message=FALSE}
ggplot(aes(x=(residual.sugar)),data=wine)+
  geom_histogram(fill="orange",binwidth = 0.04)+
  xlim(0,quantile(wine$residual.sugar,0.99))+
  scale_x_log10()
```

```{r echo=FALSE,warning=FALSE,message=FALSE}
summary(wine$residual.sugar)

```

After transfromation it can be observed that quantity of residual sugar roughly follows a bimodal distibution .the average quantity of residual sugar across the dataset is 6.391 while median value is 5.200 .There is an outlier value in the above plot which is 65.800.Lets explore it further using boxplot:

```{r echo=FALSE,warning=FALSE}
boxplot(wine$residual.sugar,horizontal = TRUE)

```

Here we can see that there are some outlier values. We will exclude them from our further analysis.Lets plot the distribution of residual sugar excluding the outliers:

```{r echo=FALSE,warning=FALSE,message=FALSE}

ggplot(aes(x=(residual.sugar))
       ,data=subset(wine,wine$residual.sugar<22))+
  geom_histogram(fill="orange",binwidth = 0.04)+
  xlim(0,quantile(wine$residual.sugar,0.99))+
  scale_x_log10()+
  xlab("Residual Sugar")


```


##4.CITRIC ACID

```{r echo=FALSE,warning=FALSE}
ggplot(aes(citric.acid),
       data=wine)+ 
  geom_histogram(fill="orange",
                 binwidth = 0.05)+
  xlim(0,quantile(wine$citric.acid,0.99))

```

```{r echo=FALSE, warning=FALSE, message=FALSE}
summary(wine$citric.acid)

```


Again ,its observed that distribution of citric acid follows normal distribution with peaking around 0.3 .the average content of Citric Acid across the dataset is 0.3342 . 

##5.VOLATILE ACIDITY

```{r echo=FALSE, warning=FALSE}
ggplot(aes(x=volatile.acidity),
       data=wine)+
  geom_histogram(fill="orange",
                 binwidth = 0.025)+
  scale_x_continuous(breaks = seq(0.1,1,0.05))
```

From the histogram we can observe that most of the values  lies under 0.6 . Lets check for outliers using boxplot:


```{r echo=FALSE,warning=FALSE,message=FALSE}
boxplot(wine$volatile.acidity,horizontal = T)


```

Here we can see that  there are few outliers . We will drop them as they may be due to some measurement errors.

```{r echo=FALSE,warning=FALSE,message=FALSE}
ggplot(aes(x=volatile.acidity),data=subset(wine,
                                           wine$volatile.acidity<0.5))+ geom_histogram(fill="orange",binwidth = 0.025)+
  scale_x_continuous(breaks = seq(0.1,0.5,0.05))+
  xlab("Volatile Acidity of the wine ")


```

```{r echo=FALSE,warning=FALSE,message=FALSE}
summary(wine$volatile.acidity)

```


The average value of volatile acidity is 0.2782 g/dm^3. Volatile Acidity represents the amount of Acetic Acid present in the wine . After removing outliers the distribution of volatile acidity follows the normal distribution with peak around 0.25 .

##6.FREE SULPHUR DIOXIDE

```{r echo=FALSE, warning=FALSE,message=FALSE}
ggplot(aes(x=free.sulfur.dioxide),data=subset(wine,
                                              wine$free.sulfur.dioxide<90))+
  geom_histogram(fill="orange",
                 binwidth = 5)+
  xlab("Amount of free sulphur dioxide in the wine ")+
  xlim(0,100)+
  scale_x_continuous(breaks = seq(0,100,5))


```

```{r echo=FALSE, warning=FALSE,message=FALSE}
summary(wine$free.sulfur.dioxide)
```

it also follows the normal distribution after removing the outliers. Average value of free sulphur dioxide across the wine dataset is 35.31 . 

##7.DENSITY 

```{r echo=FALSE,warning=FALSE}
ggplot(aes(x=density),data=wine) +
  geom_histogram(fill="orange",binwidth = 0.0005)+
  scale_x_continuous(limits = c(0.9871,1.02))


```

Here we can see that most of the wines have density falling between 0.98 and 1. We can explore outliers using boxplot as follows:

```{r echo=FALSE,warning=FALSE,message=FALSE}
boxplot(wine$density,horizontal = T)

```

We can clearly see three outliers in the distribution of density across wine quality dataset. Lets remove them and plot histogram for density variable:

```{r echo=FALSE,warning=FALSE,message=FALSE}
ggplot(aes(x=density),data=subset(wine,
                                  wine$density<1.002)) +
  geom_histogram(fill="orange",
                 binwidth = 0.0005)+
  scale_x_continuous(limits = c(0.9871,1.005))

```

```{r echo=FALSE ,warning= FALSE}
summary(wine$density)

```

Now we have a  roughly normal distribution curve for density.it can be seen that  wines have average density of 0.994

##8.pH of wine 

```{r echo=FALSE, warning=FALSE}
ggplot(aes(x=pH),data=subset(wine,
                             wine$pH<3.6 & wine$pH>2.82))+
  geom_histogram(fill="orange",
                 binwidth = 0.02)
```

```{r echo=FALSE,warning=FALSE}
summary(wine$pH)

```

Again we can see normal shape of the distribution . Most of the wines have pH around 3.2 .pH represents the sourness of the wine .Since no wine have pH >7 , none of the wine are bitter.The most acidic wine have pH of 2.720 while the least acidic wine have pH of 3.820 .

##9.CHLORIDES

```{r echo=FALSE,warning=FALSE,message=FALSE}
ggplot(aes(x=chlorides),data =subset(wine,
                                     wine$chlorides<0.075 & wine$chlorides>0.02))+geom_histogram(fill="orange",
                                                                                                 binwidth = 0.003)+
  scale_x_continuous(breaks = seq(0.02,0.08,0.01))

```

```{r echo=FALSE,warning=FALSE}
summary(wine$chlorides)

```

From summary statistics we can see that the average content of chlorides is 0.0457 g/dm^3 .After removing the outliers the plot looks more of a normal distribution.

##10.SULPHATES

```{r echo=FALSE,warning=FALSE}
ggplot(aes(x=sulphates),data=subset(wine,
                                    wine$sulphates<0.78))+
  geom_histogram(fill="orange",
                 binwidth = 0.04)+
  scale_x_continuous(breaks = seq(0.2,0.8,0.04))
```

```{r echo=FALSE,warning=FALSE}
summary(wine$sulphates)

```

We can see that the average suphate content is 0.4898 g/dm^3.The distribution of sulphates also follows the normal distribution curve.

#UNIVARIATE ANALYSIS

###Q.What is the structure of your dataset ?

There are 4986 white wines in the dataset with 12 features.The following observations can be made after doing some plottig:
1.Most of the wines have rating of 6 .
2.the median alcohol quantity is 10.40 .
3. Some wines in the dataset have  zero citric Acid .

###Q.What are the main features of interest in your dataset ?

the main feature of interest in our dataset is alcohol and quality .I would like to determine how the quantity of alcohol in the wine affects its quality .

###Q.What other features in the dataset do you think will help  support your investigation into your feature of interest ?

I think other features like pH , citirc Acid  would also affect the quality of the wine .

###Q.Of the features you investigated, were there any unusual distributions? Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?

I log transformed the skewed- distribution of residual sugar to get more even distribution of the data. After transformation bimodal curve is obtained .I also removed some outliers from the data before applying transformation.


#BIVARIATE PLOTTING

##1.CORRELATION MATRIX 

```{r echo=FALSE,warning=FALSE}
cor(wine)

```

```{r echo=FALSE, echo=FALSE, warning=FALSE}
wine<-read.csv("wineQualityWhites.csv")
library(corrplot)
corrplot(cor(wine))


```


From the coorelation matrix plot we can see that quality and alcohol have comapratively strong value of pearsons r .We will further investigae it by plotting them against each other.I have plotted some other variables of interest in the following section:

##2.QUALITY VS ALCOHOL


```{r echo=FALSE,Bivariate_Plots,warning=FALSE,message=FALSE}
wine$quality<-as.factor(wine$quality)
ggplot(aes(x=alcohol, y=quality),
       data=wine)+
  geom_point(alpha=1/5,
             position = "jitter")
  


```

As we can see that as quality increase alcohol content also increase .


##3.QUALITY VS VOLATILE ACIDITY 

```{r echo=FALSE,warning=FALSE}
ggplot(aes(x=volatile.acidity, y=quality),
       data=subset(wine,wine$volatile.acidity<0.7))+
  geom_point(alpha=1/5, position = "jitter")

```

Here we can see that wine with higher quality rating tend to have smaller volatile Acidity .


##4.QUALITY VS DENSITY 

```{r echo=FALSE, warning=FALSE}
ggplot(aes(x=density, y=quality,
           ),
       data=subset(wine,
                   wine$density<1.002))+
  geom_point(alpha=1/5, position = "jitter")


```

Again there appears to be some relationship between density and qualty as higher quality wine tend to have comparatively lower density .Although it was not our main feature of interest but after plotting the correlation matrix we saw some evidence of this relationship.

##5.QUALITY VS CHLORIDES

```{r echo=FALSE,warning=FALSE,message=FALSE}
ggplot(aes(x=chlorides,y=quality),
       data=subset(wine,wine$chlorides<0.075))+
  geom_point(alpha=1/5, position = "jitter")
```




#Bivariate Analysis

From the coorelation matrix it is evident that quality and alcohol have strong relationship as compared to other variables. This strengthened out assumption .After plotting alcohol vs quality on scatterplot we can cearly see some trends.Higher quality wines tend to have higher alcohol quantity.

I also observed some interesting relationship between qualty and density  which was not part of our initial assumption. Volatile acidity and quality also  showed some negative relationship.

The strongest relationship i found was between alcohol and quality.




#MULTIVARIATE PLOTs

##1.ALCOHOL VS VOLATILE ACIDITY AND QUALTY 

```{r echo=FALSE, Multivariate_Plots}
wine$quality <- ordered(wine$quality, levels = c(3, 4, 5, 6, 7, 8,9))

library(RColorBrewer)
myColors <- brewer.pal(7,"BrBG")
names(myColors) <- levels(wine$quality)
colScale <- scale_colour_manual(name = "quality",values = myColors)

ggplot(aes(x = alcohol, y = volatile.acidity),
       data = wine) +
  geom_point(aes(color = quality)) +
  colScale  

```


Here we can observe that wine with higer quality rating tends to have higher alcohol percentage and lesser volatile Acidity .

##2.ALCOHOL VS DENSITY AND QUALITY AS COLOR 

```{r echo=FALSE, message=FALSE, warning=FALSE}

library(RColorBrewer)
myColors <- brewer.pal(7,"Blues")
names(myColors) <- levels(wine$quality)
colScale <- scale_colour_manual(name = "quality",values = myColors)

ggplot(aes(x = alcohol, y = density),
       data =subset(wine,wine$density<1.002)) +
  geom_jitter(aes(color =factor(quality)),alpha=1/3) +
  colScale+
  geom_smooth(method=lm)


```

Here again we can see that wine with higher rating tends to have lower density and higher alcohol content.





#FINAL PLOTS AND SUMMARY 

##PLOT ONE

```{r echo=FALSE,warning=FALSE,Plot_One}

ggplot(aes(x=quality),data = wine)+geom_bar(fill="orange")+
  ggtitle("Distribution of wines across quality rating parameters")+
  xlab("Qualty (score between 0 and 10)")+
  ylab("Number of wines in dataset")
```

```{r echo=FALSE,warning=FALSE,message=FALSE}
summary(wine$quality)

```


Above plot shows the distribution of observations from dataset  across different quality parameter ranked by the experts.We can see that most of the wines have quality rating of 6.From summary statistics it is observed that maximum quality rating is 9 and minimum quality rating is 3.


##PLOT TWO

```{r echo=FALSE,warning=FALSE,message=FALSE}
ggplot(aes(x=alcohol, y=quality),data=wine)+
  geom_point(alpha=1/10, position = "jitter")+
  xlab("Alcohol (percentage by volume)")+
  ylab("Quality (score between 0 and 10)")+
  ggtitle("Quality vs Alcohol percentage ")
```

```{r echo=FALSE,warning=FALSE,message=FALSE}
by(wine$alcohol,wine$quality,summary)

```


From above plot and summary statistics we can conclude that as the alcohol percentage in wine increases its quality rating also inreases.For wines with quality rating of 3, the average alcohol content is 10.34 while for wines with quality rating of 9 , the  average alcohol content is 12.18 .In general , top quality wines contains higher alcohol content. we can infer that alcohol content is crucial factor to consider while making a good quality wine.

##2.PLOT THREE

```{r echo=FALSE,warning=FALSE,message=FALSE}
myColors <- brewer.pal(7,"Blues")
names(myColors) <- levels(wine$quality)
colScale <- scale_colour_manual(name = "quality",values = myColors)

ggplot(aes(x = alcohol, y = density),
       data =subset(wine,wine$density<1.002)) +
  geom_jitter(aes(color = factor(quality)),alpha=1/3) +
  colScale+
  ggtitle("Alcohol vs Density vs Quality")+
  xlab("Alcohol (percentage by volume)")+
  ylab("Density (g/cm^3)")+
  geom_smooth(method = lm)


```


```{r echo=FALSE,warning=FALSE,message=FALSE}
by(wine$density,wine$quality,summary)

```


Our final plot represents that higher quality rated wines have lesser density  comparatively which is clear from the summary statistics plotted above .In the plot most of higher quality wines are located in lower right corner depicting lesser density and higher alcohol content.The average density of wines with quality rating of 3 , is 0.9949 while the average density across wines with qualty rating of 9 , is 0.9915 . We can observe this trend clearly in the plot .In general , as the quality of wine is increasing , density is decreasing and the alcohol content is increasing .



#REFLECTION

From exploratory data analysis of white wine quality dataset it can be inferred that higher quality wine tends to have higher alcohol content . As per our assumption i find out the relationship between quality and alcohol content. I also found that higher quality wines have lesser density as compared to lower quality wines. So density and alcohol content is the crucial factor to consider in order to make good wines.I failed to find any other interesting relationships in the dataset due to limited flexibilty of plotting techniques and overlapping of data points across different category .It would be intersting to find out how other factors like pH , residual sugar etc affects
the quality score of a wine .If our dataset would have included informations like price of wine, manufacturer of wine etc, it would have been easy to classify wines based on quality score ratings.
In conclusion, using exploratory data analysis, I was able to see new relations, confirm the patterns already stated in the data set description, and found some clear trends in dataset.
